---
title: "2. Identifying the most reactive clonotype from donor MR4050"
author: "Friederike Dündar"
date: "July 2021; updated `r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{2. Identifying most reactive clonotype from donor MR4050}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval=TRUE, cache=TRUE
)
```
```{r libraries, message=FALSE, warning=FALSE, cache=FALSE}
library(data.table)
library(magrittr)
library(SingleCellExperiment)  
library(ggplot2); theme_set(theme_bw(base_size = 16) )
library(ggrepel)
library(patchwork)

#library(KlebanoffVDJData2020)
```


```{r cache=FALSE}
##sce.filt <- readRDS(file = paste0( "/Users/frd2007/Documents/Projects/2018-10_Smita/data/sce_", samplename, "_clonotypefiltered.rds"))
devtools::load_all()
sce.filt <- load_MR4050filt()
samplename <- "MR4050"
ifng_threshold <- 4
```


# Identifying *IFNg* threshold

```{r }
markers <- c("IFNG", "TNF", "IL2", "TNFRSF4", "TNFRSF9") 
mks.dt <- scABC2::make_long_dt(sce.filt, exprs_values = "logcounts",
  features = markers, include_metaData=c("cluster", "Sample",
    "cdr3s_aa","freq_per_Sample", "numTRA","numTRB"))
mks.dt$gene_name <- factor(mks.dt$feature_name, levels = markers, ordered = TRUE)
mks.dt <- mks.dt[logcounts > 0]
```

```{r activation_marker_expression, fig.width = 12, fig.height = 9}
ggplot(mks.dt, 
  aes(x = cluster, y = logcounts, color = freq_per_Sample)) +  
  geom_violin() +
  ggbeeswarm::geom_quasirandom(size =1) +
  facet_wrap(~gene_name) +
  facet_grid(Sample~gene_name) +
  theme(legend.position = "bottom") +
  scale_color_viridis_c()
```

```{r ifng_threshold, fig.width = 5, fig.height = 6}
ggplot(mks.dt[gene_name == "IFNG"], 
  aes(x = cluster, y = logcounts, color = freq_per_Sample)) +  
  geom_violin() +
  ggbeeswarm::geom_quasirandom(size =1) +
  facet_wrap(~gene_name) +
  geom_hline(yintercept=ifng_threshold, linetype = "dashed") +
  facet_grid(Sample~gene_name) +
  theme(legend.position = "bottom") +
  scale_color_viridis_c()
```

4 is a justifyable cut-off for IFNg expression since we're expecting a polyclonal response and I want to be a bit more lenient in the beginning.

# Identifying the most reactive clonotype

To select the best candidate, we're going to **compare the median expression of IFNg** between MUT and WT samples.

Extracting clonotypes of cells with IFNg above 4:

```{r}
cdr3s_aa_list <- unique(subset(mks.dt, gene_name == "IFNG" & logcounts > ifng_threshold)$cdr3s_aa)
```

Assess the number of cells with the respective clonotypes:

```{r clonotype_frequencies, fig.width = 12, fig.height = 9}
mks.dt[ cdr3s_aa %in% cdr3s_aa_list, .N, by = c("Sample","cdr3s_aa")] %>%
  ggplot(., aes(x = cdr3s_aa, y = N, fill = Sample)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  coord_flip() +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = c("lightskyblue1", "limegreen")) +
  ggtitle(paste("Frequencies of CDRs of cells with IFNg >", ifng_threshold),
      subtitle = samplename) +
  ylab("# cells") +
  geom_hline(yintercept = 20, linetype = "dashed")
```
 x
I don't think we need to focus on the most abundant ones here.
Smita also expects there to be a polyclonal response. 

```{r}
freq_cells <- mks.dt[ cdr3s_aa %in% cdr3s_aa_list, .N, 
    by = c("Sample","cdr3s_aa")] %>%
    #.[N>=20] %>%
    .$cdr3s_aa %>% unique
```


```{r}
strsplit(freq_cells, ";") %>% unlist %>% grep("TRA", value=TRUE, .) %>% table %>% as.data.frame
```

```{r}
strsplit(freq_cells, ";") %>% unlist %>% grep("TRB", value=TRUE, .) %>% table %>% as.data.frame
```

```{r}
med_ifng <- mks.dt[gene_name == "IFNG" & cdr3s_aa %in% freq_cells,
                   median(logcounts, na.rm = TRUE),
                   by = c("Sample","cdr3s_aa")] %>%
  dcast(.,  cdr3s_aa~Sample, value.var = "V1") 
med_ifng[is.na(MUT.MR4050), MUT.MR4050 := 0]
med_ifng[is.na(WT.MR4050), WT.MR4050 := 0]
med_ifng[, delta_MUT_WT := MUT.MR4050-WT.MR4050]
setorder(med_ifng, -delta_MUT_WT)
```

```{r table1, results='asis', cache=FALSE}
knitr::kable(as.data.frame(med_ifng))
```

```{r activation_markers_in_Ifng_high_cells, fig.height = 8, fig.width = 65, warning = FALSE}
ggplot(mks.dt[cdr3s_aa %in% freq_cells ], 
  aes(x = Sample, y = logcounts)) +  
  geom_violin(aes(color = Sample)) +
  ggbeeswarm::geom_quasirandom(size =2.5, groupOnX = TRUE, shape = 21, aes(fill = freq_per_Sample)) +
  xlab("") + coord_flip() +
  facet_grid(gene_name~factor(cdr3s_aa, levels = med_ifng$cdr3s_aa, ordered = TRUE)) +
  theme(legend.position = "bottom") +
  scale_fill_viridis_c() +
  scale_color_manual(values = c("lightskyblue1","limegreen"))
```

------------

```{r}
sessionInfo()
```

